---
title: "Reading in data and splitting by NEP"
output:
  html_notebook
---

# Introduction.
This is a quick script to split the monstrously large OA data set from multiple 
NEPs into smaller pieces, with one file for each Waterbody. Some water bodies
have two sampling locations.

# Import Packages
```{r}
library(readxl)
```

# Load Data
This is a big file, and the import can be time consuming. Providing data column
types can dramatically speed import.  If there ar any errors in the data
types, either by mis-specification, or because the automatic file parser makes
a mistake, import can slow to a crawl, as R reports import errors.

# Load Data from Other NEPs
## Establish Folder Reference
```{r}
sibfldnm <- 'Complete_NEP_Data'
parent   <- dirname(getwd())
sibling  <- file.path(parent,sibfldnm)

fn    <- 'Data_6.12.2020_v2_REG_DELIVER.xlsx'
fpath <- file.path(sibling,fn)
```

```{r}
all_data <- read_excel(fpath, 
    col_types = c("text", "text", "date", 
        "skip", "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "skip", "skip", "skip", "skip", "skip"))
```

# Split and Save by NEP
That file is so large that R has trouble with it.  Since we want to analyze by
NEP anyway (mostly) let's split them into separate files.
```{r}
for(nep in unique(all_data$WaterBody)) {
  fn <- paste0(nep, '_oa_data.csv')
  part <- all_data[all_data$WaterBody==nep,]
  write.csv(part, fn)
}

```
